# Base R vs. Tidyverse {#baser}

```{r, child='_setup.Rmd'}
```

```{r, warning = FALSE, message = FALSE}
library(tidyverse)
```

We have worked primarily with the tidyverse tools in this course. While many of the advantages of the tidyverse tools for data wrangling are their simplicity and human readability, there is actually nothing that tidyverse does which cannot be similarly achieved with baseR commands. If you are interested in comparing and contrasting check out the examples below. 

These all use the `iris` dataset, a dataset that comes with baseR (just type `iris`).


## Extract variables

```{multCode}

``{r, eval = F}
iris$Sepal.Length # single variable
iris[, c("Species", "Petal.Width")] # by name
iris[, c(5, 4)]  # by column index
``

####

``{r, eval = F}
select(iris, Species)
select(iris, Species, Petal.Width) # by name
select(iris, 5, 4)  # by column index
``

```


## Make new variables

```{multCode}

``{r, eval = F}
iris$Petal.Ratio <- iris$Petal.Length/iris$Petal.Width

iris$Sepal.Ratio <- iris$Sepal.Length/iris$Sepal.Width
``

####

``{r, eval = F}
mutate(iris, 
       Petal.Ratio = Petal.Length/Petal.Width,
       Sepal.Ratio = Sepal.Length/Sepal.Width)
``

```


Extract observations (rows)

```{multCode}

``{r}
# Using [,]
iris[iris$Petal.Width > 0.5 & iris$Species == "setosa", ]

# Using subset (works very much like dplyr::filter)
subset(iris, Petal.Width > 0.5 & Species == "setosa")
``

####

``{r}
filter(iris, Petal.Width > 0.5 & Species == "setosa")
``

```

## Arrange observations (rows)

```{multCode}

``{r, eval = F}
# descending order of species (alphabetic) followed by ascending order of Petal.Width
iris[order(rev(iris$Species), iris$Petal.Width) , ]
``

####

``{r, eval = F}
# descending order of species (alphabetic) followed by ascending order of Petal.Width
arrange(iris, desc(Species), Petal.Width) 
``

```


## Summarise observations (rows)

```{multCode}

``{r}
# Manually create a data.frame
data.frame(Petal.Length.mean = mean(iris$Petal.Length),
           Petal.Length.sd = sd(iris$Petal.Length),
           Sepal.Length.mean = mean(iris$Sepal.Length),
           Sepal.Length.sd = sd(iris$Sepal.Length))
``

####

``{r}
summarise(iris, 
          Petal.Length.mean = mean(Petal.Length),
          Petal.Length.sd = sd(Petal.Length),
          Sepal.Length.mean = mean(Sepal.Length),
          Sepal.Length.sd = sd(Sepal.Length))
``

```


## Grouped operations

```{multCode}

``{r}
# First operate in the data.frame by group (split-apply)
mtcars_by <- by(mtcars, 
   INDICES = list(mtcars$cyl, mtcars$gear),
   FUN = function(x){
     data.frame(cyl = unique(x$cyl),
                gear = unique(x$gear),
                mpg.mean = mean(x$mpg),
                mpg.sd = sd(x$mpg),
                wt.mean = mean(x$wt),
                wt.sd = sd(x$wt))
   })

# Then combine the results into a data.frame
do.call(rbind, mtcars_by)
``

####

``{r}
mtcars %>% 
  group_by(cyl, gear) %>% 
  summarise(mpg.mean = mean(mpg),
            mpg.sd = sd(mpg),
            wt.mean = mean(wt),
            wt.sd = sd(wt)) %>% 
  ungroup() # remove any groupings from downstream analysis
``

```

## Create new columns as calculations

```{multCode}

``{r, eval = F}
# First operate in the data.frame by group (split-apply)
iris_by <- by(iris, 
              INDICES = iris$Species, 
              FUN = function(x){
                x$Petal.Width.centered <- x$Petal.Width - mean(x$Petal.Width)
                return(x)
              })

# Then combine the results into a data.frame
do.call(rbind, iris_by)
``

####

``{r, eval = F}
iris %>% 
  group_by(Species) %>% 
  mutate(Petal.Width.centered = Petal.Width - mean(Petal.Width)) %>% 
  ungroup() # remove any groupings from downstream analysis
``

```


## Filter rows with conditions evaluated by group

```{multCode}

``{r}
# First operate in the data.frame by group (split-apply)
widest_petals <- by(iris, 
                    INDICES = iris$Species, 
                    FUN = function(x){
                      x[x$Petal.Width == max(x$Petal.Width), ] 
                    })

# Then combine the results into a data.frame
do.call(rbind, widest_petals)
``

####

``{r}
iris %>% 
  group_by(Species) %>% 
  filter(Petal.Width == max(Petal.Width))
``

```


## Pivot data

```{multCode}

``{r, eval = F}
reshape(iris, 
        varying = c("Sepal.Length", "Sepal.Width", "Petal.Length", "Petal.Width"),
        timevar = "trait",
        idvar = "id",
        v.names = "measurement",
        direction = "long")
``

####

``{r, eval = F}
iris %>% 
 pivot_longer(cols = Sepal.Length:Petal.Width, values_to = "measurement", names_to = "trait" )
``

```

